---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# mc.heterogeneity

<!-- badges: start -->
<!-- badges: end -->

The package mc.heterogeneity implements a Monte Carlo Based Heterogeneity Test for standardized mean differences (d), Fisher-transformed Pearson's correlations (r), and natural-logarithm-transformed odds ratio (OR) in Meta-Analysis Studies. 

Depending on the presence of moderators, this Monte Carlo Based Test can be implemented in the random or mixed-effects model. This package uses rma() function from the R package metafor to obtain parameter estimates and likelihood, so installation of R package metafor is required.

## Installation

You can install the released version of mc.heterogeneity from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("mc.heterogeneity")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
# library(devtools)
devtools::install_github("gabriellajg/mc.heterogeneity")
```

## Vignette

The vignette illustrates usage of functions in mc.heterogeneity and interpretation of the output. Type the following code to get the complete vignette:

```{r example}
library(mc.heterogeneity)
vignette("mc.heterogeneity")
```

