---
title: "A Monte Carlo Based Test for Between-study Heterogeneity in Meta-Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mc.heterogeneity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

###### 

This R package mc.heterogeneity provides functions for testing between-study heterogeneity in meta-analysis of standardized mean differences (d), Fisher-transformed Pearson's correlations (r), and natural-logarithm-transformed odds ratio (OR). 

Inclusion of moderators is an option for researchers who are interested in measuring the between-study heterogeneity per se and exploring factors that can explain the systematic between-study heterogeneity. 

In the following examples, we describe how to use our package mc.heterogeneity to test the between-study heterogeneity for each of the effect sizes considered in the current study. *Datasets*, *R codes*, and *output* are provided for each example so that applied researchers can easily replicate the examples and modify the codes for their own datasets. 

* The three example datasets are internal datasets in our package, but applied researchers can load the datasets using mc.heterogeneity:::[dataset name]. In each of the example datasets, the rows correspond to studies in meta-analysis, and the columns correspond to required input for that study, which includes, but is not limited to effect size, sample size(s), and moderators. 

* The example R codes adopt the default value for some of the arguments (e.g., default cutoff alpha level is 0.05). To change the defaults, use help() to see more details for each of the functions. 

* The output are formatted to have the same layout. 

We also include a "Empirical Illustration" section in the article to discuss the following examples. 

## 0. Installation of the package

Install the released version of mc.heterogeneity from [CRAN](https://CRAN.R-project.org) with:

```{r setup, eval = FALSE}
install.packages("mc.heterogeneity")
```

Or install the development version from [GitHub](https://github.com/gabriellajg/mc.heterogeneity) with:

```{r, eval = FALSE}
#install.packages("devtools")
library(devtools)
devtools::install_github("gabriellajg/mc.heterogeneity", force = TRUE, build_vignettes = TRUE)
library(mc.heterogeneity)
```

```{r, echo=FALSE}
library("mc.heterogeneity")
```

### 1. Standardized Mean Differences (d)

`mc.d()` is the function to test the between-study heterogeneity in meta-analysis of standardized mean differences (d). 

### 1.1 Without moderators

Load the example dataset `selfconcept` first: 

```{r}
selfconcept <- mc.heterogeneity:::selfconcept
```

Extract the required arguments from `selfconcept`:
```{r}
# n1 and n2 are lists of samples sizes in two groups
n1 <- selfconcept$n1
n2 <- selfconcept$n2
# g is a list of effect sizes
g <- selfconcept$g
```

If `g` is a list of biased estimates of standardized mean differences in the meta-analytical study, a small-sample adjustment must be applied:

```{r}
cm <- (1-3/(4*(n1+n2-2)-1)) #correct factor to compensate for small sample bias (Hedges, 1981)
d <- cm*g
```

Run the heterogeneity test using `mc.d()`:

```{r, eval=FALSE, results = 'hide'}
mc.run <- mc.d(n1, n2, est = d, model = 'random', p_cut = 0.05)
```

Alternatively, such an adjustment could be performed within the function by specifying `adjust = TRUE`:

```{r, eval=FALSE, results = 'hide'}
mc.run2 <- mc.d(n1, n2, est = g, model = 'random', adjust = TRUE, p_cut = 0.05)
```

`mc.run` and `mc.run2` will return the same results:

```{r, eval=FALSE}
mc.run
mc.run2
```

* The first line presents the results of Q-test following Restricted Maximum-Likelihood (REML) of a random-effects model. The Q-statistic is Q(df = 17) = 23.39 and the associated p-value is 0.137. Using a cutoff alpha level of either 0.05 or 0.1, this statistic is n.s (not significant). The variability in the observed standardized mean differences (d) is not larger than sampling variability alone, and the assumption of homogeneity is not rejected. 

* The second line presents the results following MC-ML-LRT. The statistic is 1.61 and the Monte-Carlo based p-value is 0.051. The assumption of homogeneity is not rejected with a cutoff alpha level of 0.05 but will be rejected at a alpha level of 0.1. 

* The third line presents the results following MC-REML-LRT. The statistic is 2.04 and the Monte-Carlo based p-value is 0.053. Similar to those following MC-ML-LR, the assumption of homogeneity is not rejected with a cutoff alpha level of 0.05 but will be rejected at a alpha level of 0.1. 


### 1.2 With moderators

Load an hypothetical dataset `hypo_moder` first: 

```{r}
hypo_moder <- mc.heterogeneity:::hypo_moder
```

Three moderators (cov.z1, cov.z2, cov.z3) are included: 

```{r}
head(hypo_moder)
```

Again, run the heterogeneity test using `mc.d()` with all moderators included in a matrix `mods` and model type specified as `model = 'mixed'`:

```{r, eval=FALSE, results = 'hide'}
mc.run3 <- mc.d(n1 = hypo_moder$n1, 
                n2 = hypo_moder$n2, 
                est = hypo_moder$d, 
                model = 'mixed', 
                mods = cbind(hypo_moder$cov.z1, hypo_moder$cov.z2, hypo_moder$cov.z3), 
                p_cut = 0.05)
```

In the presence of moderators, the results `mc.run3` will be different from those in `mc.run` and `mc.run2`:

```{r, eval=FALSE}
mc.run3
```

* The first line presents the results of Q-test following Restricted Maximum-Likelihood (REML) of a mixed-effects model in the presence of moderators. The Q-statistic is Q(df = 17) = 23.39 and the associated p-value is 0.137. Using a cutoff alpha level of either 0.05 or 0.1, this statistic is n.s (not significant). The variability in the observed standardized mean differences (d) is not larger than sampling variability alone, and the assumption of homogeneity is not rejected. 

* The second line presents the results following MC-ML-LRT. The statistic is 1.61 and the Monte-Carlo based p-value is 0.051. The assumption of homogeneity is not rejected with a cutoff alpha level of 0.05 but will be rejected at a alpha level of 0.1. 

* The third line presents the results following MC-REML-LRT. The statistic is 2.04 and the Monte-Carlo based p-value is 0.053. Similar to those following MC-ML-LR, the assumption of homogeneity is not rejected with a cutoff alpha level of 0.05 but will be rejected at a alpha level of 0.1. 


A test for (residual) heterogeneity is automatically carried out by the function. Without moderators in the model, this is simply Cochran's Q-test (Cochran, 1954), which tests whether the variability in the observed effect sizes or outcomes is larger than would be expected based on sampling variability alone. A significant test suggests that the true effects or outcomes are heterogeneous. 

When moderators are included in the model, this is the Q_E-test for residual heterogeneity, which tests whether the variability in the observed effect sizes or outcomes not accounted for by the moderators included in the model is larger than would be expected based on sampling variability alone.

#### For the following two examples (Fisher-transformed Pearson's correlations r; Natural-logarithm-transformed odds ratio OR), no moderators are included, but one can simply include moderators as in section 1.5. 

### 2. Fisher-transformed Pearson's correlations (r)

`mc.fcor()` is the function to test the between-study heterogeneity in meta-analysis of Fisher-transformed Pearson's correlations (r). 

Load the example dataset `sensation` first: 

```{r}
sensation <- mc.heterogeneity:::sensation
```

Extract the required arguments from `sensation`:

```{r}
# n is a list of samples sizes
n <- sensation$n
# Pearson's correlation
r <- sensation$r
# Fisher's Transformation
z <- 1/2*log((1+r)/(1-r))
```

Run the heterogeneity test using `mc.fcor()`:

```{r, eval=FALSE, results = 'hide'}
mc.run <- mc.fcor(n, z, model = 'random', p_cut = 0.05)
```

The test of between-study heterogeneity has the following results:

```{r, eval=FALSE}
mc.run
```

### 3. Natural-logarithm-transformed odds ratio (OR)

`mc.lnOR()` is the function to test the between-study heterogeneity in meta-analysis of Natural-logarithm-transformed odds ratio (OR). 

Load the example dataset `smoking` from R package `HSAUR2`: 

```{r}
library(HSAUR2)
data(smoking)
```

Extract the required arguments from `smoking`:

```{r}
# Y1: receive treatment; Y2: stop smoking
n_00 <- smoking$tc - smoking$qc  # not receive treatement yet not stop smoking
n_01 <- smoking$qc # not receive treatement but stop smoking
n_10 <- smoking$tt - smoking$qt # receive treatement but not stop smoking
n_11 <- smoking$qt # receive treatement and stop smoking
```

The log odds ratios can be computed, but they are not needed by `mc.lnOR()`:

```{r}
lnOR <- log(n_11*n_00/n_01/n_10)
lnOR
```

Run the heterogeneity test using `mc.lnOR()`:

```{r, eval=FALSE, results = 'hide'}
mc.run <- mc.lnOR(n_00, n_01, n_10, n_11, model = 'random', p_cut = 0.05)
```

The test of between-study heterogeneity has the following results:

```{r, eval=FALSE}
mc.run
```

